SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2012-12-24 paper/paper.tex
paper: updates.

\section{Introduction}\label{sec:intro}

The QuickCheck testing framework was a revolutionary step-forward in
type-directed random testing~\cite{qc}.  Nowadays, QuickCheck is a standard
aspect of the Haskell development cycle.  Because QuickCheck generates random
values for testing, counterexamples it finds may not be substantially larger
than a minimal counterexample.  In their original QuickCheck paper~\cite{qc},
the authors report the following user experience by Andy Gill:
%
\begin{quote}
Sometimes the counterexamples found are very large and it is difficult to go
back to the property and understand why it is a counterexample.
\end{quote}
%
\noindent
QuickCheck defines a type class \ttp{Arbitrary} that presents a method
\ttp{arbitrary} for generating random values of a given type.  Gill added another
method to the type class:
%
\begin{code}
smaller :: a -> [a] 
\end{code}
%
\noindent
The purpose of \ttp{smaller} is to generate strictly smaller values (according to
some measure) from a given counterexample to attempt to find a smaller
counterexample.  Today, \ttp{smaller} is called \ttp{shrink}.

@@

However, this straightforward approach to defining \ttp{shrink} instances may
either miss minimal counterexamples, be too inefficient to be practical, or
both.\footnote{E.g., for a recent real-world case, see the user question posed on StackOverflow:
  \url{http://stackoverflow.com/questions/8788542/how-do-i-get-good-small-shrinks-out-of-quickcheck}.}
For example, consider again Figure~\ref{fig:initial}.  The data type \ttp{B} is
a product type over lists of 16-bit signed integers, wrapped by a constructor.
Consider the property \ttp{prop} in the figure, stating that if the sum of each
list in a field of \ttp{B} is less than 16, then the sum of the four fields of
\ttp{B} is less than 64.  \ttp{prop} seems reasonable at first glance, until
one realizes that due to underflow, the property can be violated.  For example,
consider the value
%
\begin{code}
B [A (-32769)] [] [] []
\end{code}
%
\noindent
Without shrinking, QuickCheck finds large examples to the property; on average,
QuickCheck returns a value contains 70 \ttp{Int16} values!

@@

\paragraph{Contributions}
This paper, and the library it describes, makes the following contributions:

@@

To present the algorithm, consider the \ttp{SubTypes} type class below for which
QuickCheck's \ttp{Arbitrary} class is a super-class.\footnote{This presentation
  deviates from the implementation; for example, we elide details regarding
  error handling.}
%
\begin{code}
class Arbitrary a => SubTypes a where
  size    :: a -> Int
  index   :: a -> Int -> Maybe SubVal
  replace :: a -> Int -> SubVal -> a
\end{code}
%
\noindent
\ttp{SubTypes} has three methods.  The \ttp{size} method returns the size of a
value, \ttp{index} returns a sub-value at an breadth-first index in a value, and
\ttp{replace} replaces a sub-value at a particular focus.  If the index is
out-of-bounds (too large or negative), \ttp{index} return \ttp{Nothing}.  The
method \ttp{replace} returns the original value for an out-of-bounds index or an
incorrectly-typed replacement value.  The \ttp{SubVal} type is an existential
data type, used as a generic container for sub-values from a counterexample.
%
\begin{code}
data SubVal = 
  forall a. (Arbitrary a, SubTypes a) => SubVal a
\end{code}
%
\noindent
For example, consider a binary tree type
%
\begin{code}
data Tree = L | B Tree Tree
\end{code}
% 
\noindent
and the value \ttp{tree}, labeled with the indexes of its sub-values
%
\begin{code}
tree = B\sub{0} (B\sub{1} L\sub{3} 
             (B\sub{4} L\sub{6} L\sub{8}))
          (B\sub{2} L\sub{5} L\sub{7})
\end{code}
%
\noindent
Here are example applications of \ttp{SubTypes}; in the following, we show
the indexes with respect to the \ttp{tree} value:
%
\begin{code}
size tree = 9

@@

\begin{figure}[ht]
  \lstset{caption={}
%         ,xleftmargin=5ex,firstnumber=0,numbers=left
         }
  \begin{lstlisting}
reduce :: SubTypes a 
       => a -> (a -> Property) -> IO a
reduce cex prop = reduce' 1
  where
  reduce' idx =
    case index cex idx of 
      Nothing -> return cex
      Just v  -> do
        vs <- newVals v
        case test cex idx vs prop of
          Nothing   -> reduce' (idx+1)
          Just cex' -> reduce cex' prop

newVals :: SubVal -> IO [SubVal]
newVals (SubVal a) = 
  sequence (replicate maxTries s)
  where s =     sizedArbitrary a 
            >>= return . SubVal

test :: SubTypes a 
     => a -> Int -> [SubVal] 
     -> (a -> Property) -> Maybe a
test cex idx vs prop = go vs
  where
  go []      = Nothing
  go (v:vs') = 
    let cex' = replace cex idx v in
    if pass prop cex' then go vs' 
      else Just cex'
  \end{lstlisting}
  \caption{Counterexample reduction algorithm.\label{fig:reduction}}
\end{figure}

@@

For example, consider a simple calculator language containing constant
\ttp{Int}s, addition, and division:
%
\begin{code}
data Exp = C Int
         | Add Exp Exp
         | Div Exp Exp  
\end{code}
%
\noindent
and consider an interpreter for the language that returns \ttp{Nothing} if
division-by-zero is encountered and \ttp{Just} the value otherwise:
%
\begin{code}
eval :: Exp -> Maybe Int
eval (C i) = Just i
eval (Add e0 e1) =
  liftM2 (+) (eval e0) (eval e1)
eval (Div e0 e1) = 
  let e = eval e1 in 
  if e == Just 0 then Nothing 
    else liftM2 div (eval e0) e
\end{code}
%
\noindent
Suppose we restrict the set of programs with the following predicate:
%
\begin{code}
divSubTerms :: Exp -> Bool
divSubTerms (C _)         = True
divSubTerms (Div _ (C 0)) = False
divSubTerms (Add e0 e1)   =    divSubTerms e0 
                            && divSubTerms e1
divSubTerms (Div e0 e1)   =    divSubTerms e0 
                            && divSubTerms e1  
\end{code}
%
and then test the property
%
\begin{code}
div_prop e = divSubTerms e ==> eval e /= Nothing  
\end{code}
%
\noindent
The property \ttp{div\_prop} turns out to be false; for example, 
%
\begin{code}
Div (C 1) (Div (C 0) (C 1))
\end{code}
%
\noindent
is a counterexample.  In the predicate case
%
\begin{code}
divSubTerms (Div _ (C 0)) = False  
\end{code}
%
\noindent
we forgot to recur on the divisor.  However, a typical counterexample returned
by QuickCheck, without \ttp{shrink} implemented, is as follows:
%
\begin{quote}
\ttp{Div (Div (C (-22)) (Div (C 2) (C 13))) (Div (Div (Div (C (-11)) (C 14))
  (Add (Div (C (-4)) (C (-3))) (Div (C (-25)) (C (-25))))) (Add (Add (C 22) (Div
  (C (-30)) (C (-14)))) (C 13)))}
\end{quote}
%
\noindent
To butcher an expression, finding the cause of failure is like searching for a
divide-by-zero in the haystack!  But it turns out that the culprit is a small
sub-value:
%
\begin{code}
Div (C (-22)) (Div (C 2) (C 13))  
\end{code}
%

\begin{figure}
\begin{lstlisting}
reduce0 :: forall a . SubTypes a 
        => a -> (a -> Property) -> IO a
reduce0 cex prop = reduce' 1
  where
  reduce' idx = 
    case index cex idx of 
      Nothing -> return cex
      Just v  -> 
        case testHole v of
          Just v' -> reduce0 v' prop
          Nothing -> do
            vs <- newVals v
            case test cex idx vs prop of
              Nothing   -> 
                reduce' (idx+1)
              Just cex' -> 
                reduce0 cex' prop

  testHole (SubVal a) = 
    case cast a :: Maybe a of
      Nothing -> Nothing
      Just a' -> 
        if pass prop a' then Nothing
          else Just a'
\end{lstlisting}
  \caption{Reduction algorithm with the sub-value counterexample optimization.}
  \label{fig:reduce0}
\end{figure}

@@

\section{Counterexample Generalization} \label{sec:generalization}
Suppose we are testing a value.

>>>>

\section{Introduction}\label{sec:intro}
The QuickCheck testing framework was a revolutionary step-forward in
type-directed random testing~\cite{qc}.  Nowadays, QuickCheck is a standard
aspect of the Haskell development cycle.  Because QuickCheck generates random
values for testing, counterexamples it finds may not be substantially larger
than a minimal counterexample.  In their original QuickCheck paper~\cite{qc},
the authors report the following user experience by Andy Gill:
%
\begin{quote}
Sometimes the counterexamples found are very large and it is difficult to go
back to the property and understand why it is a counterexample.
\end{quote}
%
\noindent
QuickCheck defines a type class \ttp{Arbitrary} that presents a method
\ttp{arbitrary} for generating random values of a given type.  Gill added another
method to the type class:
%
\begin{code}
smaller :: a -> [a]
\end{code}
%
\noindent
The purpose of \ttp{smaller} is to generate strictly smaller values (according to
some measure) from a given counterexample to attempt to find a smaller
counterexample.  Today, \ttp{smaller} is called \ttp{shrink}.

@@

However, this straightforward approach to defining \ttp{shrink} instances may
either miss minimal counterexamples, be too inefficient to be practical, or
both.\footnote{E.g., for a recent real-world case, see the user question posed
  on StackOverflow:
  \url{http://stackoverflow.com/questions/8788542/how-do-i-get-good-small-shrinks-out-of-quickcheck}.}
For example, consider again Figure~\ref{fig:initial}.  The data type \ttp{B} is
a product type over lists of 16-bit signed integers, wrapped by a constructor.
Consider the property \ttp{prop} in the figure, stating that if the sum of each
list in a field of \ttp{B} is less than 16, then the sum of the four fields of
\ttp{B} is less than 64.  The property \ttp{prop} seems reasonable at first
glance, until one realizes that due to underflow, the property can be violated.
For example, consider the value
%
\begin{code}
B [A (-32769)] [] [] []
\end{code}
%
\noindent
Without shrinking, QuickCheck finds large examples to the property; on average,
QuickCheck returns a value contains 70 \ttp{Int16} values!

@@

\todo{check paper against this section again, especially generalization sec.}
\paragraph{Contributions}
This paper, and the library it describes, makes the following contributions:

@@

To present the algorithm, consider the \ttp{SubTypes} type class below for which
QuickCheck's \ttp{Arbitrary} class is a super-class.\footnote{This presentation
  deviates from the implementation; for example, we elide details regarding
  error handling.}
%
\begin{code}
type Idx = Int

class Arbitrary a => SubTypes a where
  size    :: a -> Idx
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
\end{code}
%
\noindent
\ttp{SubTypes} has three methods.  The \ttp{size} method returns the size of a
value, \ttp{index} returns a sub-value at an breadth-first index in a value, and
\ttp{replace} replaces a sub-value at a particular focus.  If the index is
out-of-bounds (too large or negative), \ttp{index} return \ttp{Nothing}.  The
method \ttp{replace} returns the original value for an out-of-bounds index or an
incorrectly-typed replacement value.  The \ttp{SubVal} type is an existential
data type, used as a generic container for sub-values from a counterexample.
%
\begin{code}
data SubVal =
  forall a. SubTypes a => SubVal a
\end{code}
%
\noindent
For example, consider a binary tree type
%
\begin{code}
data Tree = L | B Tree Tree
\end{code}
%
\noindent
and the value \ttp{tree}, labeled with the indexes of its sub-values
%
\begin{code}
tree = B\sub{0} (B\sub{1} L\sub{3}
             (B\sub{4} L\sub{6} L\sub{8}))
          (B\sub{2} L\sub{5} L\sub{7})
\end{code}
%
\noindent
Here are example applications of \ttp{SubTypes}; in the following, we show
the indexes with respect to the \ttp{tree} value:
%
\begin{code}
size tree = 9

@@

\begin{figure}[ht]
  \lstset{caption={}
%         ,xleftmargin=5ex,firstnumber=0,numbers=left
         }
  \begin{code}
reduce :: SubTypes a
       => a -> (a -> Property) -> IO a
reduce cex prop = reduce' 1
  where
  reduce' idx =
    case index cex idx of
      Nothing -> return cex
      Just v  -> do
        vs <- newVals v
        case test cex idx vs prop of
          Nothing   -> reduce' (idx+1)
          Just cex' -> reduce cex' prop

newVals :: SubVal -> IO [SubVal]
newVals (SubVal a) =
  sequence (replicate max s)
  where
  s = sizedArbitrary a >>= return . SubVal

test :: SubTypes a
     => a -> Idx -> [SubVal]
     -> (a -> Property) -> Maybe a
test cex idx vs prop = go vs
  where
  go []      = Nothing
  go (v:vs') =
    let cex' = replace cex idx v in
    if pass prop cex' then go vs'
      else Just cex'
  \end{code}
  \caption{Counterexample reduction algorithm.\label{fig:reduction}}
\end{figure}

@@

For example, consider a simple calculator language containing constant
\ttp{Ints}, addition, and division:
%
\begin{code}
data Exp = C Int
         | Add Exp Exp
         | Div Exp Exp
\end{code}
%
\noindent
and consider an interpreter for the language that returns \ttp{Nothing} if
division-by-zero is encountered and \ttp{Just} the value otherwise:
%
\begin{code}
eval :: Exp -> Maybe Int
eval (C i) = Just i
eval (Add e0 e1) =
  liftM2 (+) (eval e0) (eval e1)
eval (Div e0 e1) =
  let e = eval e1 in
  if e == Just 0 then Nothing
    else liftM2 div (eval e0) e
\end{code}
%
\noindent
Suppose we restrict the set of programs with the following predicate:
%
\begin{code}
divSubTerms :: Exp -> Bool
divSubTerms (C _)         = True
divSubTerms (Div _ (C 0)) = False
divSubTerms (Add e0 e1)   =    divSubTerms e0
                            && divSubTerms e1
divSubTerms (Div e0 e1)   =    divSubTerms e0
                            && divSubTerms e1
\end{code}
%
and then test the property
%
\begin{code}
div_prop e = divSubTerms e ==> eval e /= Nothing
\end{code}
%
\noindent
The property \ttp{div\_prop} turns out to be false; for example,
%
\begin{code}
Div (C 1) (Div (C 0) (C 1))
\end{code}
%
\noindent
is a counterexample.  In the predicate case
%
\begin{code}
divSubTerms (Div _ (C 0)) = False
\end{code}
%
\noindent
we forgot to recur on the divisor.

\begin{figure}[ht!]
\ttp{Div (Div (C (-22)) (Div (C 2) (C 13))) (Div (Div (Div (C (-11)) (C 14))
  (Add (Div (C (-4)) (C (-3))) (Div (C (-25)) (C (-25))))) (Add (Add (C 22) (Div
  (C (-30)) (C (-14)))) (C 13)))}
  \caption{Typical counterexample from QuickCheck with no user-defined
    \ttp{shrink} function.}
  \label{fig:div-cex}
\end{figure}

However, a typical counterexample returned by QuickCheck, without \ttp{shrink}
implemented, is shown in Figure~\ref{fig:div-cex}.  To butcher an expression,
finding the cause of failure is like searching for a divide-by-zero in the
haystack!  But it turns out that the culprit is a small sub-value:
%
\begin{code}
Div (C (-22)) (Div (C 2) (C 13))
\end{code}
%

\begin{figure}
\begin{code}
reduce0 :: forall a . SubTypes a
        => a -> (a -> Property) -> IO a
reduce0 cex prop = reduce' 1
  where
  reduce' idx =
    case index cex idx of
      Nothing -> return cex
      Just v  -> case testHole v of
                   Just v' -> reduce0 v' prop
                   Nothing -> test' v idx

  test' v idx = do
    vs <- newVals v
    case test cex idx vs prop of
      Nothing   -> reduce' (idx+1)
      Just cex' -> reduce0 cex' prop

  testHole (SubVal a) =
    case cast a :: Maybe a of
      Nothing -> Nothing
      Just a' ->
        if pass prop a' then Nothing
          else Just a'
\end{code}
  \caption{Reduction algorithm with the sub-value counterexample optimization.}
  \label{fig:reduce0}
\end{figure}

@@

\section{Counterexample Generalization} \label{sec:generalization}

Small counterexamples make debugging easier, but they are just half the battle.
To go from a specific counterexample to the required fix in a program, the
programmer must have a flash of insight in which she generalizes the
counterexample to a set of counterexamples for which the program and property
fails.  For example, in the introduction, we realized that the specific
counterexample
%
\begin{code}
B [A (-32769)] [] [] []
\end{code}
%
\noindent
is representative of the set of counterexamples in which underflow causes the
sum of the fields to be greater or equal to 64.

The generalization step is an important under-appreciated step in the debugging
process.  For counterexamples that are large, deeply-nested values, the
generalization step can be difficult.  One way we might help the programmer is
to provide her not with a single counterexample but with a quantified formula
that characterizes a set of counterexamples.  Doing so has two benefits.  First,
it already generalizes a counterexample, perhaps helping the programmer's
insight in the generalization step.  Second, it reduces the noise in favor of
the signal by abstracting away portions of large counterexample that are
irrelevant to why it violates the property.

The characterization of counterexamples that most helps the programmer seems to
should strike a middle ground.  At the one extreme, a single counterexample is
can be thought of as a closed formula containing only literals that
under-approximates the set of counterexamples.  At the other extreme, the
property itself being tested is a formula that over-approximates the failing
inputs.  We describe two kinds of formula that fall between these two extremes.

\subsection{Universal Sub-Value Generalization}
An example best explains the first kind of quantified formulas we generate.
Compare the following formula to the counterexample in Figure~\ref{fig:div-cex}:
%
\begin{code}
\ttp{forall x, y . Div (Div x (Div (C 2) (C 13))) y}
\end{code}
%
\noindent
Here, the formula claims that for all well-typed values \ttp{x} and \ttp{y}, the
property fails.  Not only does the generalization assist the programmer's
insight, but it reduces the sheer size of the counterexample, and it can do so significantly.

\begin{figure}
  \begin{code}
extrapolate :: SubTypes a
       => a -> (a -> Property) -> IO [Idx]
extrapolate cex prop = extrapolate' 1 []
  where
  extrapolate' idx idxs =
    case index cex idx of
      Nothing -> return idxs
      Just v  -> do
        vs <- extrapVals v
        if allFail cex idx vs prop
          then extrapolate' (idx+1) (idx:idxs)
          else extrapolate' (idx+1) idxs

extrapVals :: SubVal -> IO [SubVal]
extrapVals (SubVal a) =
  sequence (replicate maxExtrap s)
  where
  s = unsizedArbitrary a >>= return . SubVal

allFail :: SubTypes a
     => a -> Idx -> [SubVal]
     -> (a -> Property) -> Bool
allFail cex idx vs prop =
  (not (null res)) && and res
  where
  res  = catMaybes (map (fail prop) vals)
  vals = map (replace cex idx) vs
  \end{code}
  \caption{Universal sub-value generation algorithm.}
  \label{fig:universal}
\end{figure}

We call the kind of formula just shown \emph{universal sub-value generalization}
and is implemented in SmartCheck.  The algorithm is shown in
Figure~\ref{fig:universal}.  The algorithm is similar to the reduction algorithm
in Figure~\ref{fig:reduction} (and in the implementation, the algorithms are
generalized and combined).  The function \ttp{extrapolate} takes a
counterexample \ttp{cex} and a property.  It returns a list of indexes to be
universally quantified.  This algorithm also performs a breadth-first search
over the sub-values of the counterexample.  For each sub-value, the algorithm
generates new sub-values, replaces them in \ttp{cex}, and tests them.  If every
new counterexample non-trivially fails (i.e., does not just fail the
precondition) the property \ttp{prop}, then we extrapolate, claiming that for
\emph{any} new value replacing that sub-value in \ttp{cex}, the property will
fail.

The function \ttp{extrapVals} generates the new sub-values, much like
\ttp{newVal} in the reduction algorithm.  Here, \ttp{maxExtrap} well-typed
values are generated, and \ttp{unsizedArbitrary} is the analogue of
\ttp{sizedArbitrary}, but does not limit the size of values generated.

In the definition of \ttp{allFail},
%
\begin{code}
fail :: (a -> Property) -> a -> Maybe Bool
\end{code}
%
\noindent
is the roughly the dual of \ttp{pass} in the reduction algorithm: it returns
\ttp{Just True} if a value satisfies the property and passes the property's
precondition \ttp{pre}, \ttp{Just False} if the value fails the property, and
\ttp{Nothing} if the value fails \ttp{pre}.

Thus,

\subsection{Existential Sub-Value Generalization}

\subsection{Automated Invariant Strengthening}

\begin{figure}[ht!]
  \begin{center}
    \includegraphics[scale=0.5]{Figs/cex-gen}
  \end{center}
  \caption{Counterexample generalization.}
  \label{fig:cex-gen}
\end{figure}

\subsection{Generalization Details}

\todo{flags, optimizations}
